---
title: "Econ 448 Child Mortality Lab"
author: "Melissa Knox"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_notebook
---
## Overview and Setup
For this exercise, we will estimate the relationship between mother's education and child health, using the fraction of the mother's children who have died as our measure of child health.

These exercises use the data set "1% sample Ethiopia ever-married women age 15+" available via the Econ 448 class account from ipums.org.  Please download the data and answer the following questions.

Install these packages once (if working on your personal machine). You do not need to do this if you are working in the class server. 

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(tidyr)
library(ivreg)
library(dplyr)
library(fixest)
library(doBy)
library(stargazer)
library(coefplot)
library(ipumsr)
library(vtable)
```

Now read the data. You should have downloaded a file called "ipumsi_00007.dat.gz" from your IPUMS account. Make sure you have both files (xml and dat.gz) and that they are in the same directoy.


```{r}
ddi <- read_ipums_ddi("ipumsi_00007.xml")  
ipumsr::ipums_view(ddi)
df_ipums <- read_ipums_micro(ddi)
```

## Data Cleaning
Much of the work we do to analyze data is actually processing (or cleaning) raw data to get it ready to be analyzed.  This often means creating new variables, re-scaling existing variables, and even changing the unit of observation in each row.

```{r}
df_analysis <- as.data.frame(df_ipums)

#Filtering don't knows and nonsense values
df_analysis<-filter(df_analysis, YRSCHOOL<90)  
df_analysis<-filter(df_analysis, YRSCHOOL_SP<90) 
df_analysis<-filter(df_analysis, CHBORN<31) #Values over 30 are don't knows
df_analysis<-filter(df_analysis, CHSURV<31) #Values over 30 are don't knows
head(df_analysis)
```

We should also create some dummy variables.
```{r}
#Make dummy variables for urban, literacy, and employment
df_analysis$URBAN <- ifelse(df_analysis$URBAN == 2, 1, 0) 
df_analysis$LITERATE<-ifelse(df_analysis$LIT==2,1,0)
df_analysis$EMPLOYED<-ifelse(df_analysis$EMPSTAT==1,1,0)

#View the new variables and others of potential interest.
stargazer(df_analysis, type='text')


```



Now we need to construct a measure of child mortality, using children born and children who survived.

```{r}
#Create your new variables and then look at their summary statistics. It's always a good idea to look at your data as much as possible!
df_analysis<-transform(df_analysis,CMR=((CHBORN-CHSURV)/CHBORN))

stargazer(df_analysis[c("CHBORN","CMR")], type = "text")
```
Note that we filtered out the women with zero children when we filtered out the "don't knows" from CHSURV.



## Analysis
The first step of analysis is often visualizing our data. Lets look at a simple scatter plot.
```{r}
ggplot(df_analysis, aes(x=YRSCHOOL,y=CMR))+geom_point()
```
But it's hard to see what is happening with all of those dots, so let's try just looking at mean CMR by years of education.
```{r}
#Create a new data frame that has mean CMR by years of schooling.
#Can you look at median instead? Do you see any major differences? What about creating the same table by spouse's characteristics?
collapse <- summaryBy(CHBORN+CMR ~ YRSCHOOL, FUN=c(mean),na.rm=TRUE, data=df_analysis)
collapse
```

Now we can plot these means versus years of education.

```{r}
ggplot(collapse, aes(x=YRSCHOOL,y=CMR.mean))+geom_point()
```
Or, you could look at the same graph, but let the color of the dots reflect the mean number of children born at that level of schooling.This figure adds the extra information that women with less schooling have, on average, both more children and higher child mortality.
```{r}
ggplot(collapse, aes(x=YRSCHOOL,y=CMR.mean, color=CHBORN.mean))+geom_point()
```

An alternative to this type of plot is the "coefficient plot." With a coefficient plot, we regress the outcome on a series of dummy variables for years of education and then plot the coefficients and confidence intervals. The y-axis shows the difference between the mean outcome for each value of years of education, relative to the mean value for zero years of education.

```{r}
f_ols1<-feols(CMR~i(YRSCHOOL), data=df_analysis)
summary(f_ols1)
```

Now plot (iplot is part of the coefplot package).
```{r}
iplot(f_ols1)
```

Now let's see how the effect of schooling changes as we control for other important factors that might be associated with schooling.
```{r}
f_ols2<-feols(CMR~i(YRSCHOOL)+CHBORN+AGE+EMPLOYED+LITERATE, data=df_analysis)
iplot(f_ols2)
```
Ok, now we might want to see what we get from the control variables, so let's plot them both together. I found a package (ggiplot) that works a bit better than iplot here. Can you interpret the difference in coefficients shown in the figure below?

```{r}
install.packages("ggiplot", repos = "https://grantmcdermott.r-universe.dev")
library(ggiplot)
ggiplot(list('No Controls' = f_ols1, 'Controls' = f_ols2),main = 'Child Mortality and Education', ref.line = -1, pt.join = TRUE)

```

Just for "fun" now let's look at this by child sex.
```{r}
df_analysis<-transform(df_analysis,CMR_M=((CHBORNM-CHSURVM)/CHBORNM))
df_analysis<-transform(df_analysis,CMR_F=((CHBORNF-CHSURVF)/CHBORNF))
stargazer(df_analysis[c("CMR_M","CMR_F")], type = "text")
```

```{r}
f_ols3<-feols(CMR_M~i(YRSCHOOL), df_analysis)
f_ols4<-feols(CMR_F~i(YRSCHOOL), df_analysis)
```

```{r}
ggiplot(list('Male CMR' = f_ols3, 'Female CMR' = f_ols4),main = 'Child Survival by Sex and Education', ref.line = -1, pt.join = TRUE)
```