---
title: "Child Mortality Lab"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_notebook
author: Melissa Knox
---
## Overview and Setup
For this exercise, we will estimate the relationship between mother's education and child health, using the fraction of the mother's children who have died as our measure of child health.

These exercises use the data set "1% sample Ethiopia ever-married women age 15+" available via the Econ 448 class account from ipums.org (class code is AAP7QS).  Please download the data and answer the following questions.

```{r}
#Remember, only need to run these commands once
install.packages("ivreg", dependencies = TRUE)
install.packages("fixest")
install.packages("doBy")
install.packages("htmltools")
install.packages("shiny")
install.packages("DT")
```

Install these packages and (optional) change your working directory every time you start a fresh session.  

```{r, warning=FALSE}
library(ggplot2)
library(tidyr)
library(ivreg)
library(dplyr)
library(fixest)
library(doBy)
library(stargazer)
setwd("/Users/knoxm/Dropbox (Personal)/Pandemic Teaching/Econ 448 Data Course/")
```

Read in the data.  You will have to download it from Canvas and save it in your R working directory or in the directory you set above.

```{r}
## install.packages("ipumsr")
library(ipumsr)
##This code will read in the ipums data using the data dictionary.  When you look at your data, the variables will have labels (as will the values).
## You will also get a nice data dictionary in RStudio's viewer pane (bottom right)
ddi <- read_ipums_ddi("ipumsi_00007.xml")  ##Use whatever title your xml file has
ipumsr::ipums_view(ddi)
df_analysis <- read_ipums_micro(ddi)
```

```{r}
##Or, if you prefer to just download it yourself from ipums
df_analysis<-read.csv(file = "Ethiopia_Mothers.csv")  
```

## Data Cleaning
Much of the work we do to analyze data is actually processing (or cleaning) raw data to get it ready to be analyzed.  This often means creating new variables, re-scaling existing variables, and even changing the unit of observation in each row.

```{r}
df_analysis <- df_analysis %>% dplyr::na_if(99) #Doing this because this data uses 99 a lot for "don't know"
df_analysis<-filter(df_analysis, YRSCHOOL<90)  #Several complicated codes over 90 here, so filtering these out

df_analysis$URBAN <- ifelse(df_analysis$URBAN == 2, 1, 0) ##Creating a dummy variable that is 1 if person lives in urban area


```


The data contains a variable on total children born, but it has a lot of missing values, so I am using the sum of male children born and female children born as the measure of total children born.  I am using this information, plus the number of surviving male and female children, to create a measure of child survival ("CMR"). 

```{r}
#Create your new variables and then look at their summary statistics. It's always a good idea to look at your data as much as possible!
df_analysis<-transform(df_analysis, CHILDREN=CHBORNF+CHBORNM)
df_analysis<-transform(df_analysis, CMR=(CHBORNF+CHBORNM-CHSURVF-CHSURVM)/CHILDREN)

## But first, let's make sure we are only looking at women with children
df_analysis <- filter(df_analysis, CHILDREN>0) 
stargazer(df_analysis[c("CHILDREN","CMR")], type = "text")
```

This data set also has the following variables: years of school, literacy, and employment status.  The code below lets you look at the summary statistics for these variables.  Note that employment status is not a 0/1 dummies, but varies from 1-3.  The codebook would tell you that 1=employed, 2=unemployed, and 3=inactive. These are called factor variables. Literacy is also not 0/1, but here 1=illiterate and 2=literate.  Also note that I have included the schooling and employment values for each woman's spouse. What do you notice about the difference between schooling and employment for women and their spouses?

```{r}
stargazer(df_analysis[c("AGE", "URBAN", "RELIGION", "YRSCHOOL","YRSCHOOL_SP","LIT","EMPSTAT","EMPSTAT_SP")], type = "text")
```
You can also see from this table that employment and literacy are measured with categorical values.  Look at the data dictionary to interpret these values.

In the dictionary, you will see that LIT=1 means illiterate, while LIT=2 is literate (and 9 is unknown).  We can re-code our variables to make them easier to interpret once included in a regression,  

```{r}
##Recode to a dummy variable that is 1 if the person is literate
df_analysis$LIT<-ifelse(df_analysis$LIT==2,1,0)
summary(df_analysis$LIT)
```

Now we can look at some of these variables by years of education for the mother.  We do so by calculating the mean value of the variables for all women at each level of education.  There is a lot of missingness in the spouse's values, so I tell R to remove the NAs. We also need to first get rid of those spouses with education values from 90 to 95.
```{r}
#Create a new data frame that has mean number ofchildren, CMR, literacy, and employment status by years of schooling.
#Can you look at median instead? Do you see any major differences? What about creating the same table by literacy or employment? Or spouse's characteristics?
df_analysis <- filter(df_analysis, YRSCHOOL_SP<90)
collapse <- summaryBy(CHILDREN+CMR+LIT+EMPSTAT+YRSCHOOL_SP+EMPSTAT_SP ~ YRSCHOOL, FUN=c(mean),na.rm=TRUE, data=df_analysis)
collapse
```


```{r}
#Another way to look at the data is by plotting it.  
#Can you look at number of children by years of schooling instead?
ggplot(collapse, aes(x=YRSCHOOL,y=CMR.mean))+geom_point()
```
## Analysis
Finally, let's go back to the full sample and look at a simple linear regression of child mortality on years of school. We see that the negative relationship shown in the figure above remains and the slope is statistically different from zero.

```{r}
m_ols<-lm(CMR~YRSCHOOL, data=df_analysis)
summary(m_ols)
```
```{r}
f_ols<-feols(CMR~i(YRSCHOOL), data=df_analysis)
summary(f_ols)
```

```{r}
##install.packages("coefplot", dependencies =  TRUE)
##library(coefplot)
iplot(f_ols)
```
Now let's see how the effect of schooling changes as we control for other important factors that might be associated with schooling.
```{r}
f_ols2<-feols(CMR~i(YRSCHOOL)+AGE+URBAN, data=df_analysis)
iplot(f_ols2)
```
Ok, now we might want to see what we get from the control variables, so let's plot them both together.
```{r}
iplot(f_ols2)
iplot(f_ols, add = TRUE, shape = 25, color = "red") ##Note, the color isn't working for me, but maybe it works for you? It might be due to the color scheme. I'll keep working on it...
```

