---
title: "Data Analysis Lab 1"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
author: "Melissa Knox"
output: html_notebook
---

## Introduction to Data Analysis
In this lab, we will be practicing simple regression skills using linear regression with continuous and binary (aka "dummy") variables.

First, let's load our libraries.

```{r, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(stargazer)
```

Now, let's load our data. This data is from https://ourworldindata.org/fertility-rate#women-s-education-empirical-evidence. I cleaned it up a bit and put it in Github, but you are encouraged to explore the source.

```{r, warning=FALSE}
DF<-read.csv("https://raw.githubusercontent.com/melishka/Econ_448_DevoPop/main/womens-educational-attainment-vs-fertility.csv", header= TRUE, sep=",")
head(DF)
```

First,let's drop all of those missing values (the NAs).
```{r, warning=FALSE}
# Dropping NA values
DF <- drop_na(data=DF,Fertility)
DF <- drop_na(data=DF,Education)
head(DF)
```

It's always good practice to summarize and visualize your data before you work with and before you start describing your analysis results. Stargazer is my new favorite way to produce quick, easy to read tables.
```{r, warning=FALSE}
stargazer(DF, type="text")
```

Let's visualize the data:


```{r, warning=FALSE}
ggplot(data=DF, mapping=aes(x=Education, y=Fertility, color=Year))+geom_point()
```

Now we want to fit a line to the data:

```{r, warning=FALSE}
ols_1 <- lm(Fertility ~ Education, data=DF)
stargazer(ols_1, type="text")
```

And see what this looks like graphically:
```{r, warning=FALSE}
ggplot(data=DF, mapping=aes(x=Education, y=Fertility))+geom_point()+geom_smooth(method=lm, formula=y~x, se=TRUE)+ggtitle("Total Fertility versus Mean Years of Education for Women by Country")
```
Now, let's see if we can explain some of the variation here with some dummy variables. What seems relevant?

One thing we might want to control for is year, since we might imagine that the average number of children is declining in all countries by year.  Run the code below, is this what we find?
```{r, warning=FALSE}
DF$Year <- factor(DF$Year)
ols_2 <- lm(data=DF,Fertility~Education+Year)
stargazer(ols_2, type="text")
```
Now let's run the same regression, but only for two years of data - 1950 and 2010. Then we can use the "predict" function to plot the regression line for each of the two years. Can you describe in words what this regression is plotting? 
```{r, warning=FALSE}
# Regression with same slope but different intercepts for each year (confine to two years)
# Run in two parts so we can see the regression output first

DF_2 <- filter(DF,Year %in% c(1950,2010))

ols_3 <- lm(data=DF_2,Fertility~Education+Year)
stargazer(ols_3, type="text")
```

```{r, warning=}
DF_2 = cbind(DF_2, pred = predict(ols_3))

ggplot(data=DF_2, mapping=aes(x=Education, y=Fertility, color=Year)) + geom_point() + 
  geom_line(mapping=aes(y=pred))
```
We could also plot regression lines with different slopes and intercepts by year, but this would require a more advanced linear regression model. Luckily, ggplot can do it for us quite easily.

```{r, warning=FALSE}
# Regression lines with different slopes and different intercepts by year
# (Would require interaction between Year and Education to model in regression)

ggplot(data=DF_2, mapping=aes(x=Education, y=Fertility, color=Year, SHAPE=Year))+geom_point()+geom_smooth(method=lm, fill=NA)
```