---
title: "Data Analysis Lab 1"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
author: "Melissa Knox"
output: html_notebook
---
## Data Analysis Second Lab

In this lab, we will learn more about cleaning data and running regressions. But first, load your libraries.

```{r message=FALSE, warning=FALSE}
library(gapminder)
library(fixest)
library(ggplot2)
library(tidyverse)
library(stargazer)
library(modelsummary)
```

Now get data. We are using gapminder data and a pre-loaded dataset called "swiss." It contains province-level statistics on fertility and socio-economics in 1888 for the 47 French-speaking provinces in Switzerland. This is mid-demographic transition in this country.

```{r}
## It isn't strictly necessary to load these into data frames, but this is good practice for working with external data and it will allow you to make nice tables with stargazer
DF_gap<-as.data.frame(gapminder)
stargazer(DF_gap[c("lifeExp","gdpPercap")], type = "text", covariate.labels=c("Life Expectancy","GDP per Capita"))
```



## Linear Regression and Dummy Variables Using Gapminder


First, let's plot the gap minder data again:

```{r}
## First for all years and continents
ggplot(data=DF_gap, aes(lifeExp,log(gdpPercap), color=continent))+geom_point(alpha = 0.3,  position = position_jitter())
```

We can estimate the slope and intercept of this line

```{r, warning=FALSE}
model1<-lm(lifeExp ~ log(gdpPercap), DF_gap)
stargazer(model1, type = "text", covariate.labels=c("Log(GDP per Cap)"), dep.var.labels   = "Life Expectancy") 
```

But as you recognized in the first week exercise, this relationship is hiding some interesting heterogeneity. For example, look at the relationship between life expectancy and year:

```{r}
ggplot(data=DF_gap, aes(year, lifeExp, color=continent))+geom_point()
```

Now lets look at the original relationship for one year.

```{r}
## Then just for one year (can change year or even continent here)
gapminder %>%
  filter(year==1972) %>%
  ggplot(aes(lifeExp,log(gdpPercap), color=continent))+geom_point(alpha = 0.3,  position = position_jitter())

```

What we see in this data is that there are probably underlying patterns by year and continent that we aren't fully accounting for in a simple regression.

So let's try a model where each continent and year has its own intercept.
```{r}
model2 <- lm(lifeExp ~ log(gdpPercap) + factor(continent) + factor(year), data=DF_gap)
stargazer(model2, type="text", covariate.labels=c("Log(GDP per Cap)"), dep.var.labels   = "Life Expectancy")
```


```{r, warning=FALSE}
# Regression lines with different slopes and different intercepts by continent only

ggplot(data=DF_gap, mapping=aes(lifeExp,log(gdpPercap), color=continent, SHAPE=continent))+geom_point(alpha=0.3,position = position_jitter())+geom_smooth(method=lm, fill=NA)

```


## Regression with Control Variables Using "swiss" Data

Now let's look at data with a few more control variables.  The fertility/agriculture/catholic variables are from 0-100.
```{r}
data(swiss)
DF_swiss<-as.data.frame(swiss)
stargazer(DF_swiss, type = "text")
```
How should we look at this data on a graph?

```{r}
## Let's try to plot Fertility vs. Infant Mortality while looking at variation in different variables.
ggplot(data=DF_swiss, aes(Infant.Mortality, Fertility, color=Agriculture)) + geom_point()
```
Now let's look at these relationships a little more carefully with linear regression.
```{r, warning=FALSE}
model3 <- lm(Fertility ~ Infant.Mortality, data=DF_swiss)
stargazer(model3, type="text")
```

How do we interpret this relationship?  Does it change if we add control variables? How do you interpret the coefficients on the variables? Be sure to take statistical significance into account in your interpretation.
Note the following variable definitions:

* Fertility:‘common standardized fertility measure’
* Agriculture:	% of males involved in agriculture as occupation
* Examination:	% draftees receiving highest mark on army examination
* Education:	% education beyond primary school for draftees.
* Catholic:	% ‘catholic’ (as opposed to ‘protestant’).
* Infant.Mortality:	live births who live less than 1 year.


```{r, warning=FALSE}
model4 <- lm(Fertility ~ Infant.Mortality + Catholic + Agriculture + Education + Examination, data=DF_swiss)
stargazer(model4, type = "text")
```
Does the relationship between these variables look like what we would expect based on what we have learned and seen so far?  Why do think the differences exist?
