---
title: "Econ 448 Problem Set 1 Lab"
author: Melissa Knox
output: html_notebook  
---

These examples are meant to help you with the first problem set for Econ 448.  I am calling this the "lab" and we may go through some of it in class.
```{r message=FALSE, warning=FALSE}
## Packages you may need
library("ggplot2")
library("dplyr")
```


## Lab Exercises
```{r}
ASMR_A<-matrix(c(22,60,18),c(8,3,43), nrow=3, ncol=2, dimnames=list(c("<15","15-64",">64"), c("Pop_Percent","Deaths_per_1000")))
ASMR_A<-as.data.frame(ASMR_A)
ASMR_A

```
Multiplying two vectors together (dot product) can be used to calculate a weighted average, for example: $Y=w_{1}*x_{1}+w_{2}*x_{2}+w_{3}*x_{3}+w_{4}*x_{4}$

```{r}
## The code below produces the dot product of the two columns
## i.e. it multiplies element 1 of the first column by element 1 of the second, etc, and adds all of the results
CMR<-ASMR_A$Pop_Percent%*%ASMR_A$Deaths_per_1000/100
CMR

```

You can even use the notebook to display your results in regular text (look at the html to see this in action):

The crude mortality rate in Country A is `r CMR` deaths per 1000 people.

## Exploring Malthus
Was Malthus right? 
Read in the data.  
```{r}
malthus=read.csv(file='https://raw.githubusercontent.com/melishka/Econ_448_DevoPop/main/fertility-vs-child-mortality.csv', header=TRUE, sep=',') 
head(malthus)
```


```{r}
## Filter and plot data by year
malthus_1950<-filter(malthus, Year=="1950")
ggplot(malthus_1950, aes(x=Under_5_Mortality, y=TFR))+geom_point()
```

Just looking at this graph, do you think the relationship between TFR and under 5 mortality is positive or negative?  (See below for code that allows you to fit a polynomial to this data in case that is something you want to see.)

```{r}
malthus_1970<-filter(malthus, Year=="1970")
ggplot(malthus_1970, aes(x=Under_5_Mortality, y=TFR))+geom_point()+geom_smooth(method=lm, formula= y~poly(x,2))
```

