---
title: "R Notebook"
output: html_notebook
---

# Linear Regression Exercises with Dummy Variables and Other Controls
Preliminary set up. Install these packages.

```{r}
## For now, you need to install these every time in the cloud server.  Only once on your own machine.
install.packages("gapminder")
install.packages("fixest", dependencies=TRUE)
```

Load your libraries.

```{r message=FALSE, warning=FALSE}
## The cloud server has tidyvers and ggplot2, but be sure to install those on your home machine if you are using that.
library(gapminder)
library(fixest)
library(ggplot2)
library(tidyverse)
```

Now get data. We are using gapminder data and a pre-loaded dataset called "swiss." It contains province-level statistics on fertility and socio-economics in 1888 for the 47 French-speaking provinces in Switzerland. This is mid-demographic transition in this country.

```{r}
## It isn't strictly necessary to load these into data frames, but this is good practice for working with external data
DF_gap<-gapminder
data(swiss)
DF_swiss<-swiss

```

## Linear Regression and Dummy Variables Using Gapminder


First, let's plot the gap minder data again:

```{r}
## First for all years and continents
ggplot(data=DF_gap, aes(lifeExp,log(gdpPercap), color=continent))+geom_point(alpha = 0.3,  position = position_jitter())
```

We can estimate the slope and intercept of this line

```{r}
model1<-lm(lifeExp ~ log(gdpPercap), DF_gap)
summary(model1)
```

But as you recognized in the first week exercise, this relationship is hiding some interesting heterogeneity. For example, look at the relationship between life expectancy and year:

```{r}
ggplot(data=DF_gap, aes(year, lifeExp, color=continent))+geom_point()
```

Now lets look at the original relationship for one year.

```{r}
## Then just for one year (can change year or even continent here)
gapminder %>%
  filter(year==1972) %>%
  ggplot(aes(lifeExp,log(gdpPercap), color=continent))+geom_point(alpha = 0.3,  position = position_jitter())

```

What we see in this data is that there are probably underlying patterns by year and continent that we aren't fully accounting for in a simple regression.

So let's try a model where each country has its own intercept.
```{r}
model2 <-feols(lifeExp ~ log(gdpPercap) + factor(continent) + factor(year), data=DF_gap)
summary(model2)
```


```{r}
## We can plot something slightly different - separate intercepts and slopes for each continent:
ggplot(data=DF_gap, aes(lifeExp,log(gdpPercap), color=continent))+geom_point(alpha = 0.3,  position = position_jitter())+stat_smooth(method="lm")
```


```{r}
# Regression with same slope but different intercepts for each continent
DF_gap_72<-filter(DF_gap,year==1972)
fm <- lm(lifeExp ~ i(continent)+gdpPercap, data=DF_gap_72)
summary(fm)
DF_gap_72 = cbind(DF_gap_72, pred = predict(fm))
```

```{r}
ggplot(data=DF_gap_72, mapping=aes(x=lifeExp, y=gdpPercap, color=continent)) + 
  geom_abline(slope=fm$coefficients[2:7], intercept = fm$coefficients[1])
```

## Regression with Control Variables Using "swiss" Data

Now let's look at data with a few more control variables.  The fertility/agriculture/catholic variables are from 0-100.

```{r}
head(swiss)
```
How should we look at this data on a graph?

```{r}
## Let's try to plot Fertility vs. Infant Mortalit while looking at variation in different variables.
ggplot(data=DF_swiss, aes(Infant.Mortality, Fertility, color=Agriculture)) + geom_point()
```
Now let's look at these relationships a little more carefully with linear regression.
```{r}
model3 <- feols(Fertility ~ Infant.Mortality, data=DF_swiss)
summary(model3)
```

How do we interpret this relationship?  Does it change if we add control variables?

```{r}
model4 <- feols(Fertility ~ Infant.Mortality + Catholic + Agriculture + Education + Examination, data=DF_swiss)
summary(model4)
```
Does the relationship between these variables look like what we would expect based on what we have learned and seen so far?  Why do think the differences exist?
