---
title: "Econ 448"
subtitle: "Problem Set 1 - Due 04/17/2022 @11:59pm"
author: Melissa Knox
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_notebook  
---

This is the first problem set for Econ 448.  In it, you will answer a mix of questions that require both conceptual and analysis work.  **I prefer that you submit your entire document as one html document generated from your Markdown Notebook.** However, you may answer the concept questions in the word processor of your choice (converted to pdf), but please submit your code and results for any of the data analysis questions as a pdf or html file. You do not need to submit any of the lab exercises below.  This is just to help you learn the code for the formulas.
```{r message=FALSE, warning=FALSE}
## Packages you may need - need to load every session if you are in the cloud
## Otherwise, only load once
## Note 
install.packages("HMDHFDplus")
install.packages("tidyverse") ## You probably have this one, but just in case.

```

```{r message=FALSE, warning=FALSE}
## Load every time, but only once per session 
library("ggplot2")
library("dplyr")
library("HMDHFDplus")
```


## Lab Exercises
```{r}
##Thanks to Laura Jackson for correcting my code here!
ASMR_A<-matrix(c(22,60,18,8,3,43), nrow=3, dimnames=list(c("<15","15-64",">64"), c("Pop_Percent","Deaths_per_1000"))) 
ASMR_A<-as.data.frame(ASMR_A) 
ASMR_A

```
Multiplying two vectors together (dot product) can be used to calculate a weighted average, for example: $Y=w_{1}*x_{1}+w_{2}*x_{2}+w_{3}*x_{3}+w_{4}*x_{4}$

```{r}
## The code below produces the dot product of the two columns
## i.e. it multiplies element 1 of the first column by element 1 of the second, etc, and adds all of the results
CMR<-ASMR_A$Pop_Percent%*%ASMR_A$Deaths_per_1000/100
CMR

```

You can even use the notebook to display your results in regular text (look at the html to see this in action). This could be useful for your final paper.

The crude mortality rate in Country A is `r CMR` deaths per 1000 people.

## Exploring Malthus
Was Malthus right? Let's look at the relationship between TFR and under 5 mortality as a first step (ideally, we would have wages as well).
Read in the data from the web, or it should be in your class folder if you are on the cloud.
```{r}
malthus=read.csv(file='https://raw.githubusercontent.com/melishka/Econ_448_DevoPop/main/fertility-vs-child-mortality.csv', header=TRUE, sep=',') 
head(malthus)
```


```{r}
## Filter and plot data by year
malthus_1950<-filter(malthus, Year=="1950")
ggplot(malthus_1950, aes(x=Under_5_Mortality, y=TFR))+geom_point()
```

Just looking at this graph, do you think the relationship between TFR and under 5 mortality is positive or negative?  (See below for code that allows you to fit a polynomial to this data in case that is something you want to see.)

```{r}
malthus_1970<-filter(malthus, Year=="1970")
ggplot(malthus_1970, aes(x=Under_5_Mortality, y=TFR))+geom_point()+geom_smooth(method=lm, formula= y~poly(x,2))
```

## Questions to Submit
1. It took Finland about 50 years to go through its demographic transition around the turn of the 20th century.  For this question, go to populationpyramid.net, choose a country in Asia, Africa, or Central or South America that is currently in the third stage or later of its demographic transition.  Look back in time to find a year in which it was in the second stage.  Then figure out approximately how many years it took the country to enter the third stage  In your answer, tell me: Name of the country, year it was in the second stage, years to third stage.  Did this country take more or less time than Finland.  Try to think of at least two reasons why that might be the case and write a short description of your reasoning.

2. What is a nutritional poverty trap? Explain how a negative income shock like below average rainfall leading to decreased crop yield might lead to a poverty trap for a family? 

3. For this question, we will practice reading in some real data from the Human Mortality Database at UC Berkeley.  Before attempting this problem, visit their website at mortality.org, accept their terms of data use agreement, and get a free account.  Supply your own username and password below or, ideally, save them in a variable (i.e. _myname<-`knoxm@uw.edu') and don't include that part of the code with your submission.

Code to get to data frame called malthus for the US from 1933-2019 (all available years) is below.  **You should pick a different country, get the data as shown below, and reproduce the malthusian plot from above.** Then tell me whether you still think that the relationship is malthusian.  Or do you think it's malthusian for some years but not others? If so, which years. (Note: The list of countries is on their homepage or you can use the command getHFCcountries() to get a list.  They are mostly western countries, and I recommend picking a Scandinavian country to get a long and more interesting time series.)

```{r}
## Use the HMD package to get total population, total deaths, and total births by year.  
## Read file from web
## This gives you total male, total female, and overall total 
## deaths by year
mortality_by_year_age <- readHMDweb(CNTRY="USA", item="Deaths_1x1", username=myname, password=mypass) 

head(mortality_by_year_age)

```

This data is by age and year.  This will be useful later, but for now let's aggregate to get the total number of deaths.

```{r}
mortality_by_year <- aggregate(x=mortality_by_year_age[3:5], by=list(Year=mortality_by_year_age$Year), FUN=sum)  
mortality_by_year <- dplyr::select(mortality_by_year, Year, Total)
mortality_by_year <- dplyr::rename(mortality_by_year, Crude_Deaths=Total)  
head(mortality_by_year)
```

Now let's snag births (lucky for us, they have these as yearly totals already).
```{r}
births_by_year <- readHMDweb(CNTRY="USA", item="Births", username=myname, password=mypass) 
births_by_year <- dplyr::select(births_by_year, Year, Total)
births_by_year <- dplyr::rename(births_by_year, Crude_Births=Total) 

head(births_by_year)
```

And then get the population.  Population is also by year and age, so we need to aggregate.

```{r}
population_by_year_age <- readHMDweb(CNTRY="USA", item="Population", username=myname, password=mypass) 

## Aggregate and rename
pop_by_year <- aggregate(x=population_by_year_age[4:9], by=list(Year=population_by_year_age$Year), FUN=sum) 
pop_by_year <- dplyr::select(pop_by_year, Year, Total1) %>%  ##This symbol tells R to work on the same data frame in the next command
          dplyr::rename(Population_Total=Total1)
head(pop_by_year)
```

```{r}
## Now merge into one data frame
malthus <- merge(pop_by_year, mortality_by_year, by='Year', all=FALSE) 
malthus <- merge(malthus, births_by_year, by='Year')
malthus$CBR <- malthus$Crude_Births/malthus$Population_Total
malthus$CMR <- malthus$Crude_Deaths/malthus$Population_Total
head(malthus)

```


4. Use the same commands above to calculate and plot age specific mortality rates for one country and one year of data only. You will need the table called `Deaths_1x1' and `Population_1x1' as you used above. Note that you can use the data frames you already defined for the data by year and age if you use the same country you chose for 3, but it's not necessary.  You will additionally want to filter by year. Once you have your data, plot ASMR by age for that year.

5. Modify the code in the lab exercise above to use your findings from #4 and the population by age and year date from above to calculate the Crude Mortality Rate for that year.  Check it against what you calculated in #3 if you are using the same country.


6. Now we will calculate TFR for a country for one year. For this, you will need number of births by mother's age and number of women of that age in the population (technically, exposure to birth hazard, or person-years lived in that one year by women of that age).  We will use www.humanfertility.org.  I think you need a new account, but it's easiest if you use the same login and password for that website as you used above.  Luckily, the commands are all the same.
```{r}
## This code will get your data, but you will need to change the names of the relevant variables.
births_by_year_age <- readHFDweb(CNTRY="USA", item="birthsTR", username=myname, password=mypass)
births_by_age_1950 <- filter(births_by_year_age, Year=="1950")  ##This is one of the US's baby boom years

head(births_by_year_age)
head(births_by_age_1950)
```


Now get exposures.
```{r}
## This code will get your data, but you will need to change the names of the relevant variables.
exposure_by_year_age <- readHFDweb(CNTRY="USA", item="exposTR", username=myname, password=mypass)
exposure_by_age_1950 <- filter(births_by_year_age, Year=="1950")

head(exposure_by_year_age)
head(exposure_by_age_1950)

```

Now you can use these data to calculate a column that is age specific birth rates per year, then check the TFR formulas in your text to figure out how to turn this into TFR.

7. You are a young RA working at the World Bank.  It’s 4:55 and your boss just came into your cubicle to tell you that he has a meeting at 8 am to discuss new anti-poverty strategies in the fictional country of Portlandia.  He hands you the following information and asks for a poverty profile.  Oh, and he has a tennis game in 20 minutes, so could you include some policy recommendations and have the report on his desk by the morning?  

  The paper he hands you says:
      
      Portlandia is a small, poor country where the people are divided into four equal sized groups consisting of 
      1000 people each.  One group earns $100 a year, one earns $500 a year, one earns $900 a year, and the final 
      group earns $1500 a year.  The poverty line is set at $1000.  We have a budget of $300,000 for poverty      
      alleviation. 

    7.1 Please calculate the Head Count Ratio, and Income Gap Ratio for this country. (Include code, but this could
    just be simple calculations.)
    7.2	The World Bank’s policy is to minimize head count ratio.  What will your recommendation be?  What will the 
    new head count ratio be? (Include code)
    7.3 Do you think this is the correct approach to poverty alleviation?  Why or why not? 

