---
title: "Week 1 Exercise Econ 448"
author: "Melissa Knox"
output: html_notebook
---
## Getting Started with R
You will familiarize yourself with the R computing language, R Studio environment, and R Markdown language. If you are already familiar with these, then lucky you, you are ahead of me! If not, read on.

There are two ways to complete this project: Downloading and installing R on your own machine (or using an already installed version of R on a campus computer), or running this notebook in the cloud version of R.  

### Installing R on Your Computer
Note: RStudio seems to work better on PCs, so your mileage may vary here.

* Download [R](https://cran.r-project.org/) and [RStudio](https://www.rstudio.com/products/rstudio/download/) (recommended over using R alone)
* (Windows) Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/).  Note that there are instructions on this page for creating a text file called .Renviron that identifies a path for R to use to find Rtools.  It seemse complicated, but is fairly straightforward in practice.

### Using R Studio in the Cloud
* I have set up a JupyterLab for use in this class to make it really easy for you to run these programs in a cloud version of R Studio.  The link is in the class navigation pane.  When you open JupyterLab, you need to take one more step to get to RStudio, which is to push the blue button with the plus sign and then click RStudio.

## R Markdown and Notebooks
This is the html document produced by an [R Markdown](http://rmarkdown.rstudio.com) Notebook. You can download the code to run in R above (Code>Download Rmd). When you execute code within the notebook, the results appear beneath the code. You can download this file, experiment with it, and change it to suit your needs. When you are ready to finalize your results, make sure that you have run all of the relevant chunks of code and then click preview above to create a html document of your own with code and results included.

The grey box below is called a chunk.  Try executing this chunk by clicking the green arrow on the far left, the *Run* button within the chunk, or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*.

```{r}
# This is a comment and the # tells R not to read it.  Run the command below to print out the message "Hello World."
print ("Hello World")
```

Once you've installed RStudio, you have to install your packages (but only do this once). Open a script (File > New File > R Script) and type the commands below (or run them in the notebook).

If you use the class JupyterLab, you should not need to install any packages.
```{r}
# These packages should already be installed in the cloud version, but you may need them if you have your own installation.
# Run this code if you haven't run it before, but only needs to be run once after you have installed R.

install.packages('tidyverse', dependencies=TRUE, repos="https://cran.rstudio.com")
install.packages("rmarkdown", dependencies = TRUE)
install.packages("dplyr", dependencies = TRUE)
install.packages("xml2")
install.packages("ggplot2", dependencies = TRUE)
install.packages("gapminder", dependencies = TRUE)
install.packages("lifecycle", dependencies = TRUE)
```

After installing your packages, you need to load the ones you will be using in your program every time.  
```{r, warning=False}
# Only needs to be run once per session.

library(ggplot2)
library(gapminder)
library(dplyr)
```

Hint: You can also include the commands above at the start of a script if you decide to create and run your own code.

## Introduction to Data Work in R
 
```{r}
# Below, you can practice a calculation in R.  If you run the code below, R will output the sum.
3+1
```

```{r}
# But maybe you want to perform an operation and save the results in a variable so that you can use them later in other calculations?
# This code shows you how to assign a value to a variable x using a calculation, and then prints the value of x.
x<-3+1
x
```


```{r}
# In fact, R thinks in vectors and matrices works best when you assign a bunch of values to a variable.  
x<-c(1,2,3,4) #This makes a 1 column vector, which is like a row or column of a table.
x2<-matrix(c(1,2,3,4), nrow=2, ncol=2) #This makes a 2x2 matrix, which is like a table.
x
x2
```
You can try modifying the numbers above yourself and making different size matrices, etc.

Multiplying two vectors together (dot product) can be used to calculate a weighted average, for example: $Y=w_{1}*x_{1}+w_{2}*x_{2}+w_{3}*x_{3}+w_{4}*x_{4}$

```{r}
# Use the x vector from above and calculate the weighted average of its components if the the weights are 1/2, 1/4, 1/8, and 1/8.  The symbols %*% can be used to make the dot product.

weights<-c(0.5,0.25,0.125,0.125)
Y<-weights%*%x
Y

```


## Data Frames
Data frames are a useful way that R organizes data.  These are flexible versions of matrices that have special properties that make them easy to work with.  Later, you will import some actual data from the internet into a data frame, but for now we will use the gapminder data that is included in the package you loaded earlier.  

Gapminder (gapminder.org) data is data on life expectancy and GDP by country over time.  It can be used to calculate Preston Curves, as discussed in class.  For now, let's just perform a few operations on it.

```{r}
# Some quick summary statistics
summary(gapminder)
```

```{r}
# Load relevant data into data frame DF_2007 for the year 2007 and the continent of Africa.
# What is the median life expectancy in the continent in 2007?
DF<-data.frame(gapminder)
DF_2007<-filter(DF, year==2007 & continent=="Africa")
summary(DF_2007)
```

Plot the relationship below.  The plot represents the variation in life expectancy (within a single continent) in a single period in time.  The level of health technology should be (approximately) the same across all points and so the plotted relationship is like a snapshot of the variation in life expectancy with income at that particular time and state of the world.
```{r}
## Create a scatter plot of life expectancy and GDP per capita in 2007 for the continent of Africa.
ggplot(DF_2007, aes(x=gdpPercap, y=lifeExp, color=continent))+geom_point()

```

Now let's look at life expectancy as a function of income for a single country at different periods of time.  
```{r}
# Load relevant data into data frame DF_country for a country of your choosing for all years of data.
DF_country<-filter(gapminder, country=="Japan")
summary(DF_country)
```

Plot the relationship below.  The plot represents the evolution of life expectancy as the economy grows for one country, incorporating the social, political, and health contexts of that particular country, as well as changing health technology.
```{r}
# Create a scatter plot of life expectancy and GDP per capita for the country of Japan.  
ggplot(DF_country, aes(x=gdpPercap, y=lifeExp, color=continent))+geom_point()
```
****
## To Do
To complete this assignment, answer the following questions. 

1. Download R and RStudio and upload this .Rmd file (click the Code button at the top of the page) *or* open JupyterLab for the class and open First_Week_Exercise.Rmd there (remember to click the blue button then RStudio).
2. What is the median life expectancy in the world in 1952? In 2007?
3. Choose a country other than Japan.  Plot the life expectancy against GDP per capita from 1952 (or whatever the earliest year available is) to 2007. (Corrected from original version to remove the word median.)
4. What is your experience with R? If you haven't used R before, do you have experience with any other programming languages or statistical software (SPSS, SAS, Stata, etc.) or Excel?

You can submit these results however you choose, but a simple option would be to modify this notebook to answer the questions above, execute the relevant code, and hit "preview" in R Studio to create a html document that can be submitted with code and results.

****
