---
title: "IPUMS Data Lab"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_notebook
author: Melissa Knox
---
## Overview and Setup
For this exercise, we will practice opening, examining, and analyzing data from IPUMS, a web repository of population data.

These exercises use the data set "1% sample Botswana ever-married women age 15+" available via the Econ 448 class account from https://international.ipums.org/international/ (class code is 9JIY6C).  Please download the data and answer the following questions.

```{r}
#Remember, only need to run these commands once
pacman::p_load(fixest,ivreg,doBy,htmltools,shiny,DT, coefplot)

```

Install these packages and (optional) change your working directory every time you start a fresh session. 

```{r, warning=FALSE}
library(ggplot2)
library(tidyr)
library(ivreg)
library(dplyr)
library(fixest)
library(doBy)
library(stargazer)
library(ipumsr)
library(coefplot)
#setwd("C:/Users/knoxm/OneDrive - UW/Pandemic Teaching/Econ 448 Data Course")
```

Read in the data.  You will have to download a zipped folder called something like "Download.DAT" from the "My Data" page in IPUMS. Save this entire zipped folder as is in your R working directory or in the directory you set above. See Canvas for instructions for loading this file into the JupyterHub.

```{r}
## install.packages("ipumsr")

##This code will read in the ipums data using the data dictionary.  When you look at your data, the variables will have labels (as will the values).
## You will also get a nice data dictionary in RStudio's viewer pane (bottom right)
ddi <- read_ipums_ddi("ipumsi_00009.xml")  ##Use whatever title your file has when it downloads, but the name here should be correct for this exercise.
ipumsr::ipums_view(ddi)
df_ipums <- read_ipums_micro(ddi)  
df_analysis <-as.data.frame(df_ipums)
head(df_analysis)
```


## Data Cleaning
Much of the work we do to analyze data is actually processing (or cleaning) raw data to get it ready to be analyzed.  This often means creating new variables, re-scaling existing variables, and even changing the unit of observation in each row. Let's start by just looking at the data and the range of values. These can be cross-referenced with the data dictionary.
```{r}
stargazer(df_analysis, type = "text")
```
### Filtering Invalid Values
We see that a lot of people have values like 9 or 99 for these variables. Check the data dictionary to be sure, but these are usually codes that mean "don't know." We don't want to include these values in our analysis, so we have to figure out how to treat them. The easiest thing to do is to eliminate the individuals that have these values from our analysis data. Another option would be to set those values to N/A, but we can just filter them out for future computational ease.

```{r}
#In addition to 99, I see that some of the education variables are coded as 90, 91, etc. I will filter those out as well. 
df_analysis<-filter(df_analysis, YRSCHOOL<90)  #Several complicated codes over 90 here, so filtering these out
df_analysis<-filter(df_analysis, YRSCHOOL_SP<90)
df_analysis<-filter(df_analysis, CHBORN<31) #Values over 30 are don't knows

stargazer(df_analysis, type = "text")
```

Notice that we lost a lot of observations! However, now we have years of schooling between 0 and 18 and children born ranges between 0 and 14, which seems reasonable.

Now let's just look at a few variables of interest in a well-formatted table with summary statistics of our choosing.
```{r}

stargazer(df_analysis[c("CHBORN", "AGE", "YRSCHOOL", "YRSCHOOL_SP")], type = "text", summary.stat = c("n","mean","sd", "min", "max"), covariate.labels = c("Children Ever Born", "Age", "Years of School", "Spouse Years of School"), digits = 2, title="Summary Statistics", notes="Note: Data are 1% sample of ever-married women 15+ from 2011 Botswana census. Source: IPUMS International")
```

Also note that I have included the schooling values for each woman's spouse. What do you notice about the difference between schooling for women and their spouses?

### Coding Dummy and Categorical Variables
Many of the remaining variables in our dataset are dummy (or indicator) or categorical variables, meaning that they take on values for yes or no, or they take on a range of values that correspond to categories of answers with no intrinsic ordering of the categories (e.g., brown, blue, green for eye color). We can see this in the table below, but you will need to look at the dictionary to identify what the codes mean. 

```{r}
stargazer(df_analysis[c("ELECTRIC", "TOILET", "FLOOR", "RELIGION", "MARST", "EMPSTAT")], type = "text")
```
Some variables, like employment status may be easily recoded as dummy variables. Others, like religion, are better left as categorical variables. Using the feols command in R will allow you to turn them into a set of dummy variables in your regression models. 

Here is an example showing how to create an "EMPLOYED" dummy variable if we are only interested in difference between the employed and others.
```{r}
##Recode to a dummy variable that is 1 if the person is currently employed, and 0 otherwise. Currently EMPSTAT=2 means unemployed, =3 means inactive, and =9 means unknown/missing.
df_analysis$EMPLOYED<-ifelse(df_analysis$EMPSTAT==1,1,0)
summary(df_analysis$EMPLOYED)
```

## Visualizing and Analyzing Data
Suppose we were interested in understanding the relationship between a woman's education and her age. It might be helpful to view the average level of education for women by years of age. One way to do this is to create a new data frame that contains each age present in the data as well as the average years of schooling for all women of that age.
```{r}
#Create a new data frame that has mean years of schooling by age.

##Note: I am using pipes (the %>%) to get R to do what I want. summaryBY would be another command you could use without piping.

df_means_by_age<- df_analysis %>% 
  group_by(AGE) %>% 
  summarise(YRSCHOOL.mean=mean(YRSCHOOL)) %>%
  as.data.frame()

#Alternative code w/out pipes:
#df_means_by_age <- #summaryBy(CHILDREN+CMR+LIT+EMPSTAT+YRSCHOOL_SP+EMPSTAT_SP ~ YRSCHOOL, #FUN=c(mean),na.rm=TRUE, data=df_analysis)

head(df_means_by_age)

```


```{r}
#Another way to look at the data is by plotting it.  
#Can you look at number of children by years of schooling instead?
ggplot(df_means_by_age, aes(x=AGE,y=YRSCHOOL.mean))+geom_point()
```

### Analysis
The figure above shows a negative relationship. Let's analyze the full data set using linear regression to show the relationship between age and years of school more succinctly. 
```{r}
m_ols<-lm(YRSCHOOL~AGE, data=df_analysis)
stargazer(m_ols, type="text")
```

This result shows the negative relationship between age and years of school. A related regression can model this relationship non-linearly to show the effect of each age group on education. For this model, we use feols to turn age into a series of dummy variables (this is the i() notation).
```{r}
#Now use feols to create a series of dummy variables for each age (each coefficient + the intercept will be the average education at that age).
f_ols<-feols(YRSCHOOL~i(AGE), data=df_analysis)
summary(f_ols)
```

This regression also allows us to plot something similar to the plot above, plotting each regression coefficient and its 95% confidence interval.
```{r}
##We can actually plot each coefficient and the 95% confidence interval
iplot(f_ols)
```
Now let's see how the effect of schooling changes as we control for other important factors that might be associated with schooling.
```{r}
f_ols2<-feols(YRSCHOOL~i(AGE)+i(RELIGION), data=df_analysis)
summary(f_ols2) #Scroll to the bottom of the table to see the religion coeeficients
```

And the plot:
```{r}
iplot(f_ols2)
```
Ok, now we might want to see what we get from the control variables, so let's plot them both together.
```{r}
iplot(f_ols2)
iplot(f_ols, add = TRUE, shape = 25, color = "red") ##Note, the color isn't working for me, but maybe it works for you? It might be due to the color scheme. I'll keep working on it...
```

